# Supplementary material {#sec:supplementary-tables}

This section reports the simulated results that the figures of Section
\@ref(sec:simulation-results) do not show. All rejection rates come from
$B = 50{,}000$ replications per cell at $\alpha = 5\%$.

Figure \@ref(fig:supp-onepoint-both) repeats the design of
Section \@ref(sec:simulation-results) under the one-point alternative, in
which three groups are identical and only the fourth is shifted. All designs
are balanced and homoscedastic.

```{r supp-figure-path, include=FALSE}
## Resolved rather than hard-coded, so the child renders both through the
## parent (working directory vignettes/) and on its own (vignettes/children/).
supp_fig <- function(file) {
  cand <- file.path(c("figures", "../figures", "vignettes/figures"), file)
  hit <- cand[file.exists(cand)]
  ## Absolute: a relative path is resolved against the working directory by
  ## HTML but against the .tex location by LaTeX, and the two differ.
  if (length(hit)) normalizePath(hit[[1]], winslash = "/") else file.path("figures", file)
}
```

```{r supp-figure-captions, include=FALSE}
cap_supp_onepoint_both <- paste(
 "Power under the one-point alternative, mean shifts (0, 0, 0, 0.50),",
 "balanced homoscedastic design.",
 "(A) input distributions.",
 "(B) power of the six strategies."
)
```

```{r supp-onepoint-both, echo=FALSE, fig.cap=cap_supp_onepoint_both, out.width="85%"}
knitr::include_graphics(
 supp_fig("fleishman_4groups_power_homoscedastic_onepoint_d050.png")
)
```

## Power against the shift size

Table \@ref(tab:supp-delta-sweep) varies $\delta$ at fixed balanced sample
sizes, following the power study of Brunner et al. [@Brunner:2017, p. 1480],
where $\delta$ is the horizontal axis and $n$ is held fixed. It is the only
grid presented as a table, because its purpose is a direct comparison of
rejection rates with their Table 5. Rates are given to three decimals, which is
where the Monte Carlo error of $B = 50{,}000$ puts the last informative digit.

```{r supplement-table-helpers, echo=FALSE}
## The source tree is preferred over the installed package, because the tables
## are regenerated there. A directory only counts if it actually holds the
## table being asked for, so a stale installed copy cannot win.
supp_dir_for <- function(file) {
  cand <- c("../inst/simulations", "inst/simulations", "../../inst/simulations",
            system.file("simulations", package = "visStatistics"))
  cand <- cand[nzchar(cand)]
  hit <- cand[file.exists(file.path(cand, file))]
  if (length(hit)) hit[[1]] else ""
}

supp_table <- function(file, caption, label) {
  dir <- supp_dir_for(file)
  if (!nzchar(dir)) {
    cat("<!-- ", file, " not found; table omitted -->\n", sep = "")
    message("supplementary table missing: ", file)
    return(invisible(NULL))
  }
  d <- utils::read.csv(file.path(dir, file), check.names = FALSE,
                       stringsAsFactors = FALSE)
  n_lab <- sum(!names(d) %in% c("F", "W", "L", "KW", "SW", "SW+L"))
  if (knitr::is_latex_output()) {
    ## longtable, not table: a float cannot break across pages, and the delta
    ## sweep is 360 rows.
    al <- paste(c(rep("l", n_lab), rep("r", ncol(d) - n_lab)), collapse = "")
    hdr <- paste(names(d), collapse = " & ")
    cat("\\begin{longtable}{", al, "}\n", sep = "")
    cat("\\caption{", caption, "}\n\\label{tab:", label, "}\\\\\n", sep = "")
    cat("\\hline\n", hdr, " \\\\\n\\hline\n\\endfirsthead\n", sep = "")
    cat("\\hline\n", hdr, " \\\\\n\\hline\n\\endhead\n", sep = "")
    cat("\\hline\n\\endfoot\n")
    cat(apply(d, 1, function(r) paste(paste(r, collapse = " & "), "\\\\")),
        sep = "\n")
    cat("\n\\end{longtable}\n")
  } else {
    cat("<table>\n<caption>(#tab:", label, ") ", caption, "</caption>\n",
        sep = "")
    cat("<tr>", paste0("<th>", names(d), "</th>", collapse = ""), "</tr>\n")
    for (i in seq_len(nrow(d))) {
      cat("<tr>", paste0("<td>", unlist(d[i, ]), "</td>", collapse = ""),
          "</tr>\n")
    }
    cat("</table>\n")
  }
  invisible(NULL)
}
```

```{r supp-delta-sweep, echo=FALSE, results='asis'}
supp_table(
  "supplement_table_3_power_delta_sweep.csv",
  ## No HTML entities: "&" is LaTeX's alignment character and breaks the PDF
  ## branch. Math delimiters render in both outputs.
  paste("Power against the shift size $\\delta$ at fixed balanced $n$, SD = 1",
        "in all four groups, under the one-point alternative",
        "$\\mu = (0, 0, 0, \\delta)$ and the increasing trend alternative",
        "$\\mu = (\\delta/4, \\delta/2, 3\\delta/4, \\delta)$."),
  "supp-delta-sweep"
)
```
