---
title: "Route 1 simulation draft"
subtitle: "SW gate, SW+L gate, and Welch default"
output: html_document
bibliography: ../vignettes/REFERENCES.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

this_file <- knitr::current_input(dir = TRUE)
this_dir <- if (is.null(this_file)) getwd() else dirname(this_file)
root <- normalizePath(file.path(this_dir, ".."), mustWork = TRUE)

A}

shape_name <- function(distribution, gamma_shape) {
  if (distribution[1] == "normal") {
    "normal, mean = 0, SD = 1"
  } else {
    sprintf(
      "standardised Gamma, α = %s",
      format(gamma_shape[1], trim = TRUE, scientific = FALSE)
    )
  }
}

skew_name <- function(distribution, skew, gamma_shape) {
  if (distribution[1] == "normal") {
    "0"
  } else {
    format(skew[1], trim = TRUE, scientific = FALSE)
  }
}

excess_kurtosis_name <- function(distribution, excess_kurtosis, gamma_shape) {
  if (distribution[1] == "normal") {
    "0"
  } else {
    format(excess_kurtosis[1], trim = TRUE, scientific = FALSE)
  }
}

make_type_table <- function(dat, include_kw = TRUE) {
  pieces <- split(dat, dat$skew)
  out <- lapply(pieces, function(s) {
    row <- data.frame(
      Distribution = shape_name(s$distribution, s$gamma_shape),
      Skewness = skew_name(s$distribution, s$skew, s$gamma_shape),
      `Excess kurtosis` = excess_kurtosis_name(
        s$distribution,
        s$excess_kurtosis,
        s$gamma_shape
      ),
      F = range_pct(s$fisher_rejection),
      W = range_pct(s$welch_rejection),
      L = range_pct(s$levene_route_rejection),
      SW = range_pct(s$sw_rejection),
      `SW+L` = range_pct(s$sw_gate_rejection),
      `SW to W` = range_pct(s$sw_route_welch_probability),
      `SW to KW` = range_pct(s$sw_route_rank_probability),
      `SW+L to F` = range_pct(s$route_fisher_probability),
      `SW+L to W` = range_pct(s$route_welch_probability),
      `SW+L to KW` = range_pct(s$route_rank_probability),
      check.names = FALSE
    )
    if (include_kw) {
      row <- cbind(
        row[1:6],
        KW = range_pct(s$rank_rejection),
        row[7:ncol(row)]
      )
    }
    row
  })
  do.call(rbind, out)
}

make_power_table <- function(dat) {
  pieces <- split(dat, dat$skew)
  out <- lapply(pieces, function(s) {
    best <- pmax(
      s$fisher_power,
      s$welch_power,
      s$mean_power,
      s$rank_power,
      s$sw_power,
      s$gate_power
    )
    data.frame(
      Distribution = shape_name(s$distribution, s$gamma_shape),
      Skewness = skew_name(s$distribution, s$skew, s$gamma_shape),
      `Excess kurtosis` = excess_kurtosis_name(
        s$distribution,
        s$excess_kurtosis,
        s$gamma_shape
      ),
      F = pct(mean(s$fisher_power)),
      W = pct(mean(s$welch_power)),
      L = pct(mean(s$mean_power)),
      KW = pct(mean(s$rank_power)),
      SW = pct(mean(s$sw_power)),
      `SW+L` = pct(mean(s$gate_power)),
      `SW+L regret` = pct(mean(best - s$gate_power), digits = 2),
      `SW to W` = range_pct(s$sw_route_welch_probability),
      `SW to KW` = range_pct(s$sw_route_rank_probability),
      `SW+L to F` = range_pct(s$route_fisher_probability),
      `SW+L to W` = range_pct(s$route_welch_probability),
      `SW+L to KW` = range_pct(s$route_rank_probability),
      check.names = FALSE
    )
  })
  do.call(rbind, out)
}

identical_type1 <- subset(
  type1,
  design %in% c("balanced n, equal SD", "unbalanced n, equal SD")
)

unequal_type1 <- subset(
  type1,
  !design %in% c("balanced n, equal SD", "unbalanced n, equal SD")
)

mc_se <- function(p, B = 50000) sqrt(p * (1 - p) / B)
```

## Route 1 decision problem

Abbreviations used throughout: `F` = Fisher's ANOVA; `W` = Welch's
heteroscedastic ANOVA; `L` = Levene gate selecting F or W; `KW` =
Kruskal--Wallis; `SW` = Shapiro--Wilk gate selecting W or KW; `SW+L` =
Shapiro--Wilk gate selecting KW or, if the mean branch is retained, L.

Route 1 is the only branch in which `visstat()` lets assumption tests select the
final test. The current code fits a group model, tests the standardised
residuals, and uses the Shapiro--Wilk result as the first gate: rejected
residual normality routes to Wilcoxon or Kruskal--Wallis; retained residual
normality keeps the result in the mean-based branch. The design question is
therefore specific: should the parametric side of the SW gate default directly
to Welch (`SW`), or should it include the Levene gate (`SW+L`) and select
Fisher/Student only when homoscedasticity is not rejected?

Rochon et al. [@Rochon:2012] is used only as background here: two-stage testing
can change conditional error rates, so the final procedure must be judged by its
simulated final-test rejection rates.

The argument is read as figure-table pairs. Figure 1/Table 1 checks whether the
candidate routes keep the nominal rejection rate when both the equal-means and
rank-distribution nulls are true. Figure 2/Table 2 keeps the equal-means null
true but violates equality of distributions, so F, W, and L remain mean-null
tests while SW and SW+L become final-test rejection rates after possible
hypothesis switching. Figure 3/Table 3 then asks the power question under real
ordered group differences, and the final Fisher/Welch table isolates the
parametric branch.

## Simulation design

The simulation layout follows the competitor logic. Blanca et al.
[@Blanca:2017] assess Fisher's ANOVA under zero group effect, non-normal
distributions, balanced and unbalanced group sizes, and Bradley's robustness
criterion. Delacre et al. [@Delacre:2019] cross sample-size ratios, SD ratios,
and sample-size/SD pairings to compare Fisher-type and Welch-type ANOVA.
Fagerland and Sandvik [@Fagerland:2009] separate the equal-means target from
rank-based targets and show why rank tests must not be read as mean tests when
distributional shapes differ. We therefore use two Type I settings: identical
distributions, where both the equal-means and equal-rank-distribution nulls are
true, and equal means with unequal distributions, where only the equal-means
null is the common target for F, W, and L.

The normal input is the standard normal distribution,
\(X \sim \mathcal{N}(0,1)\): mean 0, variance 1, skewness 0,
and kurtosis 3. Excess kurtosis is kurtosis minus 3, so it is 0 for the
standard normal distribution. Gamma input follows the standardised Gamma
path. For
\(X \sim \Gamma(\alpha,\theta)\), with shape parameter
\(\alpha>0\) and scale parameter \(\theta>0\), the mean is
\(E(X)=\alpha\theta\), the variance is
\(\mathrm{Var}(X)=\alpha\theta^2\), the skewness is \(2/\sqrt{\alpha}\),
and the kurtosis is \(3+6/\alpha\). The excess kurtosis, reported in the
tables, is therefore \(6/\alpha\)
[@Forbes:2010, p. 109]. In the
simulations \(\theta=1\), \(\alpha\in\{16,4,1,0.25\}\), and the draws are
standardised by \(Z=(X-\alpha\theta)/(\theta\sqrt{\alpha})\). Thus the Gamma
columns deliberately vary skewness and excess kurtosis together.

All simulation cells use \(B=50{,}000\) Monte Carlo replications. For an
estimated rejection probability \(\hat p\), where \(\hat p\) is the rejection
rate shown in the corresponding cell, and for gates the rejection rate of the
final selected test,
\[
SE_{\mathrm{MC}}=\sqrt{\hat p(1-\hat p)/B}.
\]
This is about `r pct(mc_se(0.05), 2)` percentage points at
\(\hat p=0.05\). The maximum is `r pct(mc_se(0.50), 2)` percentage points at
\(\hat p=0.50\).

## Identical distributions and identical means

Figure \@ref(fig:route1-identical-typeI) is the clean null case. All groups have
the same population mean and the same distribution; therefore the equal-means
null is true for F, W, and L, and the rank-distribution null is also true for
KW. This is the only Type I figure where standalone KW is a valid comparator.

```{r route1-identical-typeI, out.width="100%", fig.cap="Route 1 Type I simulation under identical distributions and identical means. The heatmap reports final-test rejection rates at α = 5%. Green marks Bradley's 2.5%--7.5% interval. Cells for SW and SW+L additionally report the route split."}
knitr::include_graphics(
  figure_path("route1_identical_distributions_typeI_with_kw_skew4_B50000.png"),
  rel_path = FALSE
)
```

Table \@ref(tab:route1-identical-table) summarises the same figure by
distribution shape. Across the identical-distribution cells, SW and SW+L remain
inside Bradley's interval. Their mean final-test rejection rates are
`r pct(mean(identical_type1$sw_rejection), 2)` and
`r pct(mean(identical_type1$sw_gate_rejection), 2)`, respectively. The
important point is not that SW fails to route to KW under skew; it does route to
KW. The point is that this does not create false rejection here, because the
rank-distribution null is true as well.

```{r route1-identical-table}
grouped_table(
  make_type_table(identical_type1, include_kw = TRUE),
  caption = "Figure 1 summary by distribution shape. Entries are ranges across the displayed n-regimes.",
  id_cols = c("Distribution", "Skewness", "Excess kurtosis"),
  reject_cols = c("F", "W", "L", "KW", "SW", "SW+L"),
  route_cols = c("SW to W", "SW to KW", "SW+L to F", "SW+L to W", "SW+L to KW"),
  reject_header = "α-error / rejection rate of final test",
  route_header = "Routing rate: test selected"
)
```

## Equal means but unequal distributions

Figure \@ref(fig:route1-unequal-typeI) is the mean-null stress test. The group
means are equal, but the group distributions differ through the SD design.
Therefore F, W, and L still assess the equal-means null, but KW is not a
standalone Type I comparator for that null. SW and SW+L are retained in the
figure because they are the actual automatic procedures; their cells report the
rejection rate of the final selected test after a possible switch to KW.

<!-- ============================================================
REMOVE LATER WHEN THIS FIGURE IS MOVED INTO A PAPER PANEL:
this chunk currently prints the long standalone caption
============================================================ -->
```{r route1-unequal-typeI, out.width="100%", fig.cap=route1_unequal_caption}
knitr::include_graphics(
  figure_path("route1_equal_means_unequal_distributions_without_kw_skew4_B50000.png"),
  rel_path = FALSE
)
```

Table \@ref(tab:route1-unequal-table) is the numerical companion to Figure
\@ref(fig:route1-unequal-typeI). Read strictly as equal-means testing, Welch is
the most stable of the mean-based strategies in this simulation:
the mean rejection rates are F = `r pct(mean(unequal_type1$fisher_rejection), 2)`,
W = `r pct(mean(unequal_type1$welch_rejection), 2)`, and
L = `r pct(mean(unequal_type1$levene_route_rejection), 2)`. SW and SW+L are much
higher, `r pct(mean(unequal_type1$sw_rejection), 2)` and
`r pct(mean(unequal_type1$sw_gate_rejection), 2)`, because the SW gate often
switches to KW. This is the central interpretation: SW and SW+L are method
switches, not better equal-means tests.

```{r route1-unequal-table}
grouped_table(
  make_type_table(unequal_type1, include_kw = FALSE),
  caption = "Figure 2 summary by distribution shape. Entries are ranges across the unequal-distribution n/SD regimes.",
  id_cols = c("Distribution", "Skewness", "Excess kurtosis"),
  reject_cols = c("F", "W", "L", "SW", "SW+L"),
  route_cols = c("SW to W", "SW to KW", "SW+L to F", "SW+L to W", "SW+L to KW"),
  reject_header = "Mean-null rejection rate for F/W/L; final-test rejection for SW/SW+L",
  route_header = "Routing rate: test selected"
)
```

## Power under ordered group differences

Figure \@ref(fig:route1-power) addresses the complementary question: if real
ordered group differences exist, how much power is lost or gained by SW and
SW+L? This completed power simulation is balanced and homoscedastic, with four
group means shifted by 0, 0.25, 0.50, and 0.75 SD. It is therefore a power
simulation, not a Type I simulation, and it does not replace Figure
\@ref(fig:route1-unequal-typeI).

```{r route1-power-pdf, out.width="100%", fig.cap="Theoretical input PDFs for the Route 1 power simulation. Groups A--D use the same standardised normal or Gamma shape, with additive mean shifts of 0, 0.25, 0.50, and 0.75 theoretical standard deviations. Dashed vertical lines mark the group means. The density scale is clipped at 1.1 for readability in the high-skew Gamma panel."}
knitr::include_graphics(
  figure_path("gamma_skew_sweep_4groups_power_pdf.png"),
  rel_path = FALSE
)
```

```{r route1-power, out.width="100%", fig.cap="Power simulation for four balanced homoscedastic groups with ordered additive mean shifts of 0, 0.25, 0.50, and 0.75 SD. The y-axis is the probability that the final test rejects at α = 5%. The bottom annotations show the route probabilities for the SW+L gate."}
knitr::include_graphics(
  figure_path("gamma_skew_sweep_4groups_power.png"),
  rel_path = FALSE
)
```

Table \@ref(tab:route1-power-table) summarises Figure \@ref(fig:route1-power)
by distribution shape; the figure keeps the n-specific detail. Around normality
and mild Gamma skew, SW+L has slightly higher power than SW because the Levene
gate sometimes retains Fisher in homoscedastic data. At stronger skew, both SW
and SW+L route almost entirely to KW and their power is practically identical.

```{r route1-power-table}
grouped_table(
  make_power_table(power),
  caption = "Figure 3 summary by distribution shape. Power entries are means across n = 10, 20, 50, 100, and 200; the figure shows the n-specific values.",
  id_cols = c("Distribution", "Skewness", "Excess kurtosis"),
  reject_cols = c("F", "W", "L", "KW", "SW", "SW+L", "SW+L regret"),
  route_cols = c("SW to W", "SW to KW", "SW+L to F", "SW+L to W", "SW+L to KW"),
  reject_header = "Power / rejection rate of final test",
  route_header = "Routing rate: test selected"
)
```

## Fisher versus Welch inside the parametric branch

The power simulation behaves as expected in the balanced homoscedastic and balanced  corner:
The relation between Fisher's  $F$ statistic and the Welch Onewway ANOVA $F_W$ 

\[
\frac{F}{F_W}=1+\frac{3}{5(n-1)}.
\]
Thus \(F/F_W-1=3/[5(n-1)]\), so the statistic-level relative difference is
\(O(1/n)\). Note that this is not a power-ratio identity; the observed rejection rates
only show the expected ordering 
But this small homoscedastic advantage is not an argument for variance pretesting.
The literature consistently separates the two issues: Welch-type tests are
recommended as unconditional mean tests when equal variances are doubtful
[@Welch:1951; @Rasch:2011; @Delacre:2019], whereas choosing Student/Fisher or
Welch after a preliminary variance test is discouraged because the selection
step changes the operating characteristics of the final test
[@Moser:1992; @Zimmerman:2004; @Rochon:2012].

The only remaining reason for `SW+L` is therefore not statistical optimality for
the equal-means null, but consistency with the displayed model. Route 1 shows
the residual diagnostics of the equal-variance Gaussian group model; if both
displayed GLM assumptions pass, keeping Fisher/Student available makes the
reported test correspond to the model just diagnosed. If that design link is
not wanted, the literature and the present simulations both favour the simpler
`SW` route: Shapiro--Wilk retained means Welch, Shapiro--Wilk rejected means KW.

```{r route1-parametric-power, out.width="100%", fig.cap="Parametric-branch power under normality, balanced groups, and true homoscedasticity."}
knitr::include_graphics(
  figure_path("gamma_skew_sweep_4groups_parametric_branch_power.png"),
  rel_path = FALSE
)
```

```{r route1-parametric-table}
normal_power <- subset(power, distribution == "normal")
normal_power <- normal_power[order(normal_power$n_per_group), ]

parametric_table <- data.frame(
  n = normal_power$n_per_group,
  `F/F_W theory` = sprintf(
    "%.4f",
    1 + 3 / (5 * (normal_power$n_per_group - 1))
  ),
  `F power` = pct(normal_power$fisher_power, 1),
  `W power` = pct(normal_power$welch_power, 1),
  `L power` = pct(normal_power$mean_power, 1),
  `SW+L to F` = pct(normal_power$route_fisher_probability, 1),
  `SW+L to W` = pct(normal_power$route_welch_probability, 1),
  `SW+L to KW` = pct(normal_power$route_rank_probability, 1),
  check.names = FALSE
)

knitr::kable(
  parametric_table,
  caption = "Fisher/Welch comparison in the normal homoscedastic corner of the power simulation.",
  align = "r",
  escape = FALSE
)
```

## Interpretation for Route 1

If Route 1 is read strictly as an equal-means machine, the simulations support
Welch as the clean default. That is the Delacre/Rasch/Zimmerman conclusion and
is visible in Figure \@ref(fig:route1-unequal-typeI), where W is more stable
than F or L when means are equal but distributions differ. If Route 1 is allowed
to use residual shape as a method switch, SW is the simpler automatic route.
SW+L is a model-display compromise: it preserves the Fisher/Student GLM branch
when both residual normality and homoscedasticity are retained, but its
alpha-error cost must be read from the Type I simulations, not from the
Fisher/Welch power comparison.

Thus the simulations do not prove that SW+L is the best equal-means procedure.
They show the price and benefit of the automatic route. The price is that, under
equal means but unequal distributions, SW and SW+L often report the rejection
rate of KW after the hypothesis has changed. The benefit is that the switch is
visible and diagnostic-driven. The final design decision is therefore not
statistical optimality in the abstract, but which default the package wants to
make explicit: Welch for fixed equal-means inference, SW for a parsimonious
shape-based switch, or SW+L for the same switch while preserving the GLM
Fisher/Student branch when its displayed assumptions pass.

For later reference, the automatic Route 1 default is SW+L, not because it is
the statistically optimal equal-means procedure, but because it is the only
default consistent with the graphical contract of `visstat()`: the software
shows residual normality and homoscedasticity diagnostics for the Gaussian
equal-variance GLM, so if both displayed assumptions pass, the output may remain
in the Fisher/Student branch. The literature still supports Welch as the safer
unconditional mean-test default when the only goal is equal-means inference and
variance pretesting is discouraged [@Welch:1951; @Rasch:2011; @Zimmerman:2004;
@Delacre:2019], and this is visible in Figure
\@ref(fig:route1-unequal-typeI). The simulations show that SW+L behaves close to
SW while preserving the Fisher/Student branch when both displayed assumptions
pass (Figures \@ref(fig:route1-identical-typeI)--\@ref(fig:route1-power)). The
honest interface is therefore: automatic = SW+L as a transparent
assumption-based teaching/default route; parametric override = Welch for fixed
equal-means inference; nonparametric override = Wilcoxon/Kruskal--Wallis when
the user wants the rank-distribution comparison.
