---
title: "Why the central-tendency gate rests on the CLT, not on residual normality"
subtitle: "Reference note for visStatistics (not part of the package vignette)"
author: "Sabine Schilling"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6.5, dpi = 110,
  fig.path = "figures-clt-gate/"          # PNGs persist here for later reference
)
```

> **Purpose.** A personal record of the argument behind `visstat()`'s
> central-tendency gate, with the runnable code and the output figures in one
> place. The runnable script is
> [`dev/clt_mean_demo.R`](clt_mean_demo.R); the Type I companion is
> [`inst/appendix_typeI_simulation.R`](../inst/appendix_typeI_simulation.R).
> This file is a reference note, **not** the package vignette.

# The claim

The residual-normality test (Shapiro–Wilk / Anderson–Darling) inspects the
**wrong object**. Inference on group means in the general linear model — the
estimated coefficients $\hat\beta$ — relies on the **sampling distribution of
the mean** being normal, not on the residuals being normal. The central limit
theorem (CLT) normalises the former and never the latter when the
data-generating distribution is skewed or heavy-tailed.

## The argument, in the author's words

1. tests of glm family assume normal errors
2. if the function they come from is highly skewed, the residuals will never be
   normal "enough" -> fall back to non-parametric methods
3. role of clt: it assumes that when the group numbers go to infinity, the
   DISTRIBUTION of the mean, in the glm the estimated bs become normal
4. For simulation studies one must sample out of a skewed function and plot the
   distribution of the means, not the residuals, they should be normal
5. the threshold will always be dependent of the exact distribution shape,
   described by skew and kurtosis roughly. thus every threshold is arbitrary!!!

## The same chain, expanded

1. General linear model inference (the $t$/$F$ reference distributions) is
   exact when the **errors** $\varepsilon_i$ are normal. Residuals
   $e_i = y_i-\hat y_i$ only estimate the errors.
2. If the data come from a highly skewed function, the errors are skewed at
   **every** $n$; the residual sample mirrors that, so SW/AD reject — correctly
   — and routing falls back to a rank-based test.
3. The CLT acts on the **estimator**, not the errors:
   $\hat\beta=(X^\top X)^{-1}X^\top y$ is a linear combination of the $y_i$, so
   its sampling distribution $\to$ normal as $n\to\infty$ even while the errors
   stay skewed. For a one-way design $\hat\beta$ is just the group means.
4. The right thing to inspect is therefore the sampling distribution of the
   **mean**, not the residuals. That is what the simulations below plot.
5. The $n$ at which the mean is "normal enough" depends on the exact
   distribution shape (captured to leading order by skewness and kurtosis), so
   any fixed threshold is necessarily approximate.

# The law: skew of the mean $\sim \gamma_1/\sqrt n$

This is not one named theorem. It is two **exact** moment identities, plus the
expansion that uses them.

## Exact part — cumulant additivity

For iid draws, cumulants of a sum add: $\kappa_r(S_n)=n\,\kappa_r(X)$ for
$S_n=\sum_i X_i$. The mean is $\bar X = S_n/n$, so
$$\kappa_r(\bar X)=\frac{\kappa_r(S_n)}{n^{\,r}}=\frac{\kappa_r(X)}{n^{\,r-1}}.$$
Substituting into skewness $\kappa_3/\kappa_2^{3/2}$ and excess kurtosis
$\kappa_4/\kappa_2^{2}$:
$$\operatorname{skew}(\bar X)=\frac{\kappa_3/n^2}{(\sigma^2/n)^{3/2}}=\frac{\gamma_1}{\sqrt n},
\qquad
\operatorname{exkurt}(\bar X)=\frac{\kappa_4/n^3}{(\sigma^2/n)^{2}}=\frac{\gamma_2}{n}.$$
These are **exact**, not asymptotic. Averaging $n$ iid values divides the third
cumulant by an extra $n$ and the fourth by an extra $n^2$, while the variance
drops only by $n$ — hence skewness fades as $1/\sqrt n$ and excess kurtosis a
full order faster, as $1/n$.

## Edgeworth expansion — the shape

The same cumulants give the shape of the standardised mean
$Z_n=\sqrt n(\bar X-\mu)/\sigma$:
$$f_{Z_n}(x)\approx\phi(x)\Big[1+\tfrac{\gamma_1}{6\sqrt n}He_3(x)
+\tfrac{\gamma_2}{24\,n}He_4(x)+\dots\Big].$$
The $1/\sqrt n$ on the skew term and $1/n$ on the kurtosis term are exactly the
cumulant orders above.

## Berry–Esseen — the rigorous rate

$$\sup_x\bigl|F_{Z_n}(x)-\Phi(x)\bigr|\le \frac{C\,\rho}{\sigma^3\sqrt n},
\qquad \rho=E|X-\mu|^3.$$
Guarantees the CLT for any finite-third-moment distribution and pins the rate
at $1/\sqrt n$, with the leading constant scaling with skewness.

**Consequence.** A fixed per-group threshold can never be right for all shapes:
the $n$ you need scales with $\gamma_1^2$. Income-level skew ($\gamma_1\approx6$)
needs $n\gtrsim 3600$ to get the mean's skew under $0.1$.

# Simulations

Monte Carlo **illustrates** specific distributions; it is not a proof (a proof
would need every finite-moment distribution). The exact moment law above is the
general statement. Each table below should show the empirical skew/kurtosis of
the mean sitting on the predicted $\gamma_1/\sqrt n$ and $\gamma_2/n$.

```{r load-defs}
## Load the machinery from the runnable script WITHOUT triggering its auto-run.
options(clt_demo_defs_only = TRUE)
source("clt_mean_demo.R")
```

## A. Heavy tails, modest skew — the mean normalises

Johnson SU curve with skewness 2 and excess kurtosis $\approx$ 22. Despite the
enormous kurtosis, the mean is near-normal by $n=50$–$100$, because kurtosis
decays as $1/n$.

```{r scenario-johnson}
invisible(do.call(run_clt_demo, SCENARIOS$johnson))
```

## B. Income (lognormal) — extreme skew, the mean does NOT normalise

Lognormal$(0,1)$, the textbook income shape, skewness $\approx 6$. The mean's
skew is $\gamma_1/\sqrt{50}\approx0.85$ at $n=50$ and still $\approx0.2$ at
$n=1000$ — visibly non-normal across the whole 50–100 range.

```{r scenario-income}
invisible(do.call(run_clt_demo, SCENARIOS$income))
```

# Takeaways

- **Skew is the binding term.** It fades only as $1/\sqrt n$; kurtosis fades as
  $1/n$. So "kurtosis matters too" is true (it is in the expansion) but it is
  second-order and tamed quickly — case A confirms it.
- **Extreme skew defeats a fixed gate.** Income-like data (case B) keeps a
  visibly skewed mean at $n=50$–$100$; the mean-based test there carries that
  skew into its sampling distribution and inflates Type I error. This matches
  the Type I sweep in
  [`inst/appendix_typeI_simulation.R`](../inst/appendix_typeI_simulation.R),
  where Bradley's bound breaks once skewness reaches $\approx 6$ at
  $n\in(50,100)$.
- **The residual test is the wrong instrument** in this regime: it can only ever
  tell you the residuals are non-normal (they always are, for skewed data),
  which is not the question. The question is whether the *mean's* distribution
  is normal — a property of $n$ and shape, governed by the law above.

# Why no hard gate exists

It is tempting to invert the law into a skewness-aware gate. Fix a tolerance
$\delta$ on the mean's skew; then
$$\operatorname{skew}(\bar X)=\frac{\gamma_1}{\sqrt n}\le\delta
\;\Longleftrightarrow\;
n\ge\Big(\frac{\gamma_1}{\delta}\Big)^2
\;\Longleftrightarrow\;
\gamma_{1,\max}(n)=\delta\sqrt n.$$

Calibrated against the Type I sweep — using the **smallest** group in each
design as the binding $n$ — a single constant separates every cell perfectly:
the test stays within Bradley's band while the binding group's mean-skew is
below $\approx 0.6$ and inflates once it crosses it.

```{r delta-calibration, echo=TRUE}
typeI <- matrix(c(
  0.0515,0.0540,0.0540, 0.0575,0.0555,0.0525, 0.0675,0.0595,0.0525,
  0.0780,0.0680,0.0560, 0.0830,0.0700,0.0595, 0.0905,0.0760,0.0625,
  0.0975,0.0805,0.0650), byrow = TRUE, ncol = 3,
  dimnames = list(skew = 2:8, regime = c("(50,100)", "(>100)", "(>500)")))
n_min <- c(55, 110, 510)                 # smallest group per regime
sm    <- outer(2:8, n_min, function(a, b) a / sqrt(b))   # skew of the mean
over  <- typeI > 0.075                    # breaches Bradley's upper bound
c(max_skewmean_ok   = max(sm[!over]),
  min_skewmean_over = min(sm[over]))      # the two never overlap -> delta ~ 0.6
```

So the gate could read $\hat\gamma_1 \le 0.6\sqrt{n_{\min}}$. But this **runs in
circles** — it trades one arbitrary input for two:

1. **$\delta$ is empirical.** $\delta\approx0.6$ is fitted to *this* sweep
   ($k=5$, this heteroscedasticity, Welch). The link "skew of the mean $\to$
   Type I of the joint test" is not a cumulant identity, so the constant moves
   with the design.
2. **$\gamma_1$ is empirical, and unknowable here.** Applying the rule needs the
   *population* skewness; you only have the sample $\hat\gamma_1$ from the same
   group. That estimate is downward-biased for skewed data and hard-capped at
   $\hat\gamma_1\le(n-2)/\sqrt{n-1}$ (≈7.2 at $n=50$). For income the true
   lognormal skewness is 6.18 while even a 2,000,000-draw sample read 5.99 — a
   single group of 50 cannot see the tail that defines the skewness.

The CLT is asymptotic; it names no finite cutoff. Any finite gate — flat
$n>50$ or skewness-aware — is a pragmatic heuristic, not a derivable truth.
This is the Lumley/Fagerland/Franc position: a normality (or skewness)
pre-test cannot be a principled hard gatekeeper. The flat $n>50$ is not *more*
arbitrary than the skewness rule; it is *less fragile*, because it does not
depend on estimating $\gamma_1$ from a small skewed sample. That is the
argument for keeping the simple gate — while stating in the prose that it is a
convention, not a theorem.

# What the gate actually means — stated plainly

The gate does not detect "the right test." It chooses a **question** on the
user's behalf, because the choice between the mean and the median is a property
of the question, not of the data (Section
[Why no hard gate exists]). So it must be written without euphemism:

> **Above 50 observations per group, `visstat()` always compares means —
> however skewed the distribution.** The residual-normality test is not even
> consulted in this regime. This is a deliberate convention, not an inference:
> the central limit theorem makes inference on the *mean* valid at this $n$ for
> most shapes, so the tool defaults to reporting the mean. It does **not** claim
> the mean is the right summary of your data. For heavily skewed data the mean
> and the median answer different questions (e.g. total vs. typical income); if
> the median or stochastic dominance is what you care about, the mean-based
> result above the gate is answering a question you did not ask.

Two consequences a reader must not miss:

- The gate's promise is about the **validity of mean inference** (Type I error
  near nominal), nothing more. It is silent on whether the mean is *meaningful*.
- A non-significant mean comparison above the gate does **not** imply the groups
  are alike: their medians and spreads can differ sharply while their means
  coincide (the equal-means / unequal-medians construction). "Means do not
  differ" is the entire claim — no more.

# Files

| File | Role |
|---|---|
| [`dev/clt_mean_demo.R`](clt_mean_demo.R) | Runnable script; `source()` it to draw both figures to the active device. |
| `dev/clt_gate_notes.Rmd` | This reference note. |
| `dev/figures-clt-gate/` | Saved PNG figures from the chunks above. |
| [`inst/appendix_typeI_simulation.R`](../inst/appendix_typeI_simulation.R) | Type I error sweep (downstream consequence of the same mechanism). |

```{r session, echo=FALSE}
sessionInfo()
```
