# Route 1 simulations {#sec:simulation-results}

Route 1 can answer two different questions. Mean-based tests — Student's t-test,
Fisher's ANOVA, and their Welch variants — test equality of population means
[@Welch:1951; @Rasch:2011; @Delacre:2019]. Rank-based tests — the Wilcoxon rank-sum
test and the Kruskal--Wallis test — test a rank-based distributional target that
coincides with a location shift only under additional shape assumptions
[@Fagerland:2009]. `visstat()` commits to one of the two on every call, and the
simulations below quantify what that commitment costs for each of the three
implemented Route 1 options.

The first option, `group_test = "welch"`, remains mean-based and uses Welch's t-test
for two groups and Welch's one-way ANOVA for more (`W`). A Welch default avoids the
Type I error introduced by preliminary variance tests
[@Moser:1992; @Zimmerman:2004; @Rochon:2012] and gives directly interpretable
estimates and confidence intervals [@Harrell:2026], but it can be weak for strongly
skewed small samples [@Blanca:2017; @Fagerland:2012] and can answer the wrong
scientific question when the mean lies in a long tail
[@Fagerland:2009; @Harrell:2026]. The cost of defaulting to Welch-type mean
comparisons is small in the equal-variance balanced case: Welch's t-test then
coincides with Student's t-test, and for more than two balanced groups the Welch ANOVA
statistic converges to Fisher's statistic with a relative difference of order $n^{-1}$
(Eqs. \@ref(eq:welch-student-pooled) and \@ref(eq:welch-anova-equal-var-reduction)).

The Central Limit Theorem (CLT) allows the formal assessment of residual normality
assumed in Welch tests to be bypassed in large samples: by the CLT, the sampling
distribution of the group means approaches normality as the group sizes grow,
regardless of the shape of the underlying distribution, so that in large enough
samples a rejected residual-normality test no longer necessarily threatens the
validity of the mean-based routes, Fisher/Student and Welch alike. This raises the
practical question of what should count as a "large enough" sample. The answer depends
on the shape of the underlying distribution, especially skewness and tail weight, here
parametrised by excess kurtosis [@Lumley:2002], and the simulations below give it for
the shapes and designs simulated.

The second option, `group_test = "rank"`, remains rank-based and uses `wilcox.test()`
for two groups and `kruskal.test()` for more (`KW`). This reduces distributional
assumptions and can be efficient under skew
[@Bridge:1999; @Harrell:2026; @Franc:2025], but changes the estimand
[@Fagerland:2009] and gives less direct mean-scale interpretation [@Harrell:2026].

The third option is the default `group_test = NULL`, which lets residual shape decide.
The Shapiro--Wilk test on the standardised residuals of the fitted general linear
model routes rejected residual normality to `wilcox.test()` or `kruskal.test()`, and
retained residual normality to the mean-based branch, where the Levene test then
selects between the equal-variance tests (Student's t-test, Fisher's ANOVA; `F`) and
the Welch variants (`SW+L`). Test selection based on assumption tests suffers from the
fact that all hypothesis tests are underpowered in smaller samples, and thus often
fail to detect assumption violations, and overpowered in large samples
[@Ghasemi:2012; @Fagerland:2012; @Shatz:2024], so that in larger samples even mildly
skewed data are sent to the rank-based route.

The simulations display the fixed tests and the routed procedures side by side: fixed
Fisher/Student (`F`), fixed Welch (`W`), a Levene gate between them (`L`), fixed
Kruskal--Wallis (`KW`), a Shapiro--Wilk gate between `W` and `KW` (`SW`), and the full
Shapiro--Wilk plus Levene gate (`SW+L`), which selects `F` or `W` within the
mean-based branch and `KW` otherwise. Of these, `W`, `KW` and `SW+L` are the three
implemented options; `F`, `L` and `SW` are references. `L` isolates the variance gate
for mean comparison and `SW` isolates the residual-shape gate, so the difference
between `SW` and `SW+L` measures what the second gate contributes.

The aim of the simulations is not to re-establish the robustness evidence for Fisher's
ANOVA under homoscedasticity and non-normality [@Blanca:2017] or for Welch's ANOVA
under heterogeneous variances [@Delacre:2019]. Instead they show under which
combinations of sample size, skewness and excess kurtosis the displayed residual
diagnostics route from the mean-based branch to rank-based tests, when the Levene gate
keeps Fisher/Student or selects a Welch variant, and where each strategy leaves the
robustness bounds used below.

All simulations compare four groups, so every reported rate belongs to the ANOVA
member of each family; the two-group members follow the same routing with `t.test()`
and `wilcox.test()` in place of `aov()`, `oneway.test()` and `kruskal.test()`. The two
Type I designs follow published settings: Figure \@ref(fig:route1-identical-typeI)
adapts the three-group zero-effect one-way ANOVA setting of Blanca et al.
[@Blanca:2017] to four Route 1 groups with identical distributions, and
Figure \@ref(fig:route1-unequal-typeI) varies the group standard deviations and the
sample-size pairings as in Delacre et al. [@Delacre:2019].

Each cell uses $B = 50{,}000$ Monte Carlo replications. For an estimated rejection
probability $\hat{p}$ the Monte Carlo standard error is
$SE_{\mathrm{MC}}=\sqrt{\hat{p}(1-\hat{p})/B}$ [@Koehler:2009], about 0.10 percentage
points at $\hat{p}=0.05$ and at most 0.22 percentage points at $\hat{p}=0.50$.
Non-normal data are generated with Fleishman's polynomial transformation
$Y = a + bX + cX^2 + dX^3$, where $X \sim N(0,1)$ and the coefficients set mean 0,
variance 1, and the target skewness and excess kurtosis [@Fleishman:1978]. Five input
distributions are used throughout, with (skewness, excess kurtosis) equal to
$(0,0)$, $(0,6)$, $(0.5,1)$, $(1,1.6)$ and $(2,6.6)$; the first two are symmetric.
Balanced designs use $\mathbf n=(n,n,n,n)$ with $n \in \{10,20,50,100\}$; unbalanced
designs use $\mathbf n=\bar n\cdot(0.5,0.8,1.2,1.5)$, giving group sizes
$(5,8,12,15)$, $(10,16,24,30)$, $(25,40,60,75)$ and $(50,80,120,150)$ for
$\bar n \in \{10,20,50,100\}$. Equal-SD designs use $\mathbf s=(1.0,1.0,1.0,1.0)$;
unequal-SD designs use $\mathbf s=(1.0,1.3,1.7,2.2)$, and in the unbalanced case this
vector is either kept, so that larger groups carry larger SDs, or reversed to
$\mathbf s=(2.2,1.7,1.3,1.0)$, so that larger groups carry smaller SDs.

An empirical rate is reported as non-robust when it leaves Bradley's liberal bounds of
$[2.5\%, 7.5\%]$ at $\alpha = 5\%$ [@Bradley:1978]. This is a reporting convention
without principled status — Bradley's own argument is that no fixed interval defines
robustness — and it is used here only so that the same threshold is applied in every
figure. In Figure \@ref(fig:route1-identical-typeI) the colour scale is capped at
7.5%, so any exceedance would be visible at a glance; none occurs.

Three rules apply when reading the heatmaps.

1. Rows `F`, `W` and `L` always report the rejection rate of a mean-based test, so
   whenever the population means are equal they are Type I error rates in every cell.
2. Rows `KW`, `SW` and `SW+L` report the rate of whichever test was finally selected.
   They are Type I error rates as long as the rank-based null is true as well, which
   is the case when the group distributions differ at most by a symmetric rescaling.
   Under skewness combined with SD scaling the rank-based null is false, and the
   rank-routed share of such a cell is a rejection rate against a different
   hypothesis, not a Type I error rate.
3. The percentages after the vertical bar are route probabilities, not rejection
   rates: `F` and `W` for the `L` row, `W` and `KW` for the `SW` row, and `F`, `W` and
   `KW` for the `SW+L` row. They identify which test produced the rate to their left.

## Type I error simulations: equal means, equal ranks, balanced or unbalanced group sizes

Figure \@ref(fig:route1-identical-typeI) shows in panel (A) the five input
distributions, which the four groups share; the groups differ only in size. Panel (B)
varies the common group size $n \in \{10,20,50,100\}$ from top to bottom, panel (C)
the target mean group size $\bar n$ of the unbalanced design over the same values.
Because all four groups are identically distributed, the equal-means null and the
rank-based null are simultaneously true, and by rule 2 every row of both heatmaps is a
Type I error rate. This is therefore the clean Type I check of the automated routes:
a switch to the rank-based branch cannot change the truth status of the hypothesis
tested.

```{r simulation-captions, include=FALSE}
cap_route1_identical_typeI <- paste(
 "Route 1 Type I simulation under identical distributions and identical",
 "means, with group mean 0 and SD 1 in all four groups.",
 "(A) input distributions.",
 "(B) balanced design with group sizes $n_1, n_2, n_3, n_4 = n$ ",
 "listed from top to bottom as 10, 20, 50, 100. ",
 "(C) Unbalanced design with group sizes $n_1, n_2, n_3, n_4 = \\bar{n} \\cdot ",
 "(0.5, 0.8, 1.2, 1.5)$ with $\\bar{n} \\in \\{10, 20, 50, 100\\}$ ",
 "listed from top to bottom as 10, 20, 50, 100, rounded up to the next integer. The heatmaps in",
 "(B) and (C) report final-test rejection rates at $\\alpha = 5\\%$.",
 " Cells for SW and SW+L",
 "additionally report the route split. In (A), dashed lines",
 "mark means and dotted lines mark medians. The index i = 1, ..., 4",
 "denotes groups; $\\bar n$ denotes the target mean group size for unbalanced",
 "designs. All heatmap numbers are percentages; the first value is the",
 "final-test rejection rate, and gated rows list route splits after |."
)

cap_route1_unequal_typeI <- paste(
 "Route 1 simulation: identical means and unequal distributions;",
 "$\\mathrm{mean}_i = 0$ ($i = 1, \\ldots, 4$).",
 "Route 1 equal-means simulation with varied group standard deviations",
 "and sample-size pairings. (A) input distributions. (B) balanced design,",
 "listed from top to bottom as 10, 20, 50, 100.",
 "(C) unbalanced design with larger $n_1, n_2, n_3, n_4$ paired with larger SD,",
 "with rows listed from top to bottom as 10, 20, 50, 100.",
 "(D) unbalanced design with larger $n_1, n_2, n_3, n_4$ paired with smaller SD,",
 "with rows listed from top to bottom as 10, 20, 50, 100.",
 "F, W, and L are mean-null tests; SW and SW+L report final-test",
 "rejection after possible routing to Kruskal--Wallis. In (A), dashed",
 "lines mark means and dotted lines mark medians. The",
 "index i = 1, ..., 4 denotes groups; $\\bar n$ denotes the target mean",
 "group size for unbalanced designs, listed from top to bottom as 10, 20, 50, 100. All heatmap numbers are percentages;",
 "the first value is the final-test rejection rate, and gated rows list",
 "route splits after |."
)
```

```{r route1-identical-typeI, echo=FALSE, fig.cap=cap_route1_identical_typeI, out.width="95%"}
knitr::include_graphics(
 "figures/route1_identical_distributions_typeI_with_kw_fleishman_B50000.png"
)
```

No cell of either heatmap leaves Bradley's bounds. Across both designs, all five input
distributions and all four sample sizes, `F` ranges from 4.43% to 5.18%, `L` from
4.32% to 5.33%, `KW` from 4.28% to 5.06%, `SW` from 4.24% to 5.57% and `SW+L` from
4.53% to 5.71%. The widest spread belongs to fixed Welch, from 3.63% in the balanced
design at $n=10$ for the symmetric heavy-tailed input to 6.80% in the unbalanced
design at $\bar n=10$ and $\bar n=20$ for the most skewed input. The robustness of
Fisher's ANOVA to non-normality under homoscedasticity established by Blanca et al.
[@Blanca:2017] therefore extends, in this grid, to both Welch variants, to the rank
test and to both gated procedures; adding the Levene gate to `SW` never moves the rate
by more than 0.92 percentage points.

This is also where the price of the Welch default can be read against its algebraic
size. For four balanced homoscedastic groups, $F/F_W = 1+3/[5(n-1)]$
(Eq. \@ref(eq:welch-anova-four-groups)), so the Fisher and Welch statistics differ by
6.7% at $n=10$ and by 0.6% at $n=100$. In the normal column of panel (B) the
corresponding rejection rates differ by at most 0.06 percentage points at every $n$:
a difference of several per cent in the test statistic is invisible in the decision it
produces.

The route probabilities show how the two gates behave when their assumptions hold and
when they are mildly violated. Under the normal input the Shapiro--Wilk gate sends
4.67%, 5.01%, 4.85% and 5.06% of replications to the rank branch at $n=10, 20, 50$ and
$100$ — it does not drift with sample size — and `SW+L` selects Fisher's ANOVA in
89.01% to 90.09% of replications throughout. Under mild skewness the same gate drifts
strongly: for the $(0.5,1)$ input the rank route rises from 16.95% at $n=10$ to
31.79%, 64.12% and 90.24%, and for the $(2,6.6)$ input it is 87.57% at $n=10$ and
100% from $n=50$ on. The Levene gate drifts in the same direction. This is the
underpowered-in-small-samples and overpowered-in-large-samples behaviour of assumption
tests, visible here as route probabilities rather than as an argument from the
literature: in large samples even a mild shape departure is routed to the rank branch,
at a Type I cost that is nil in this design.

## Equal means, unequal ranks

In Figure \@ref(fig:route1-unequal-typeI) all four group means remain zero, but the
common standardised input distribution is multiplied by group-specific standard
deviation scale factors. The balanced block (panel B) uses $\mathbf n=(n,n,n,n)$ with
$\mathbf s=(1.0,1.3,1.7,2.2)$. The unbalanced blocks use
$\mathbf n=\bar n\cdot(0.5,0.8,1.2,1.5)$ paired either with
$\mathbf s=(1.0,1.3,1.7,2.2)$, so that larger groups carry larger SDs (panel C), or
with $\mathbf s=(2.2,1.7,1.3,1.0)$, so that larger groups carry smaller SDs (panel D).

The equal-means null is true in all five columns, so by rule 1 the rows `F`, `W` and
`L` are Type I error rates throughout. For the two symmetric columns the rescaling
leaves group means, medians and mean ranks equal, so the rank-based null is true as
well and rows `SW` and `SW+L` are Type I error rates there too. In the three skewed
columns the rescaling shifts group medians and mean ranks, the rank-based null becomes
false, and by rule 2 only the mean-routed share of a `SW` or `SW+L` cell retains a
Type I interpretation. No fixed `KW` row is shown, because a fixed rank test has no
equal-means Type I interpretation once skewed distributions are rescaled.

```{r route1-unequal-typeI, echo=FALSE, fig.cap=cap_route1_unequal_typeI, out.width="85%"}
knitr::include_graphics(
 "figures/route1_equal_means_unequal_distributions_fleishman_B50000.png"
)
```

One route probability in this figure needs explanation, because it is otherwise
paradoxical: in the normal column, where every group is exactly normal, the
Shapiro--Wilk gate still routes 13.31%, 22.78%, 45.04% and 72.47% of replications to
the rank branch in panel (D). The gate is applied to the standardised residuals of the
common fitted model, which divides all groups by a single residual standard error. If
the group variances differ, the pooled residuals are a scale mixture of normal
distributions and therefore leptokurtic, so residual normality is correctly rejected
with increasing frequency as $n$ grows even though no group is non-normal. Under
heteroscedasticity the residual-shape gate thus responds to the variance
heterogeneity, not only to the shape of the input distributions.

## Where Bradley's bounds are exceeded

Two of the five designs produce no exceedance at all: both equal-SD designs of
Figure \@ref(fig:route1-identical-typeI), as reported above, and the balanced
unequal-SD design of panel (B), where `F` reaches at most 6.19%, `W` at most 6.91% and
`L` at most 6.76%, in each case at $n=10$ under the $(2,6.6)$ input.
Heteroscedasticity alone, at equal group sizes, is thus not enough to make any of the
six strategies non-robust in this grid, which is the numerical counterpart of the
recommendation to prefer balanced designs [@Delacre:2019].

Every exceedance comes from an unbalanced design, and its direction is set by which
groups carry the large SD. Under equal means and unequal variances,
$E[MS_\text{within}] = \sum_i (n_i-1)\sigma_i^2/(N-k)$ weights each group variance by
its own degrees of freedom, whereas
$E[MS_\text{between}] = \sum_i \sigma_i^2(1-n_i/N)/(k-1)$ weights it by $1-n_i/N$.
Assigning the large SD to the small group therefore enlarges the numerator and
shrinks the denominator of Fisher's $F$, and the reverse assignment does the opposite;
both weights are ratios, so the distortion does not diminish as the sample sizes grow
[@Glass:1972]. Panel (D) is the first case: `F` rejects between 11.84% and 13.11% at
every input distribution and at every $\bar n$ from 10 to 100, with no tendency to
decrease. Panel (C) is the reverse: `F` is conservative throughout, between 2.55% and
3.11%, and falls below Bradley's lower bound at $\bar n=10$ and $\bar n=20$ for the
symmetric heavy-tailed input, at 2.40% and 2.38%. Fisher's ANOVA and Student's t-test
are therefore not weakly but systematically affected by the direction of the imbalance
under unequal variances, in both directions.

Fixed Welch leaves Bradley's bounds in only two cells of the entire grid, both in
panel (D) and both at the most extreme input $(2,6.6)$: 10.83% at $\bar n=10$ and
9.09% at $\bar n=20$. At $\bar n=50$ it is back inside at 7.22%, and at $\bar n=100$
at 6.26%. Delacre et al. [@Delacre:2019] established that the empirical Type I error
of Welch's one-way ANOVA — whose two-group special case is Welch's t-test — remained
within Bradley's bounds across their simulated non-normal distributions, which
included skewed chi-square cases with skewness 2 and excess kurtosis 6, close to the
most extreme input used here, and a symmetric mixed-normal case in which 90% of the
observations are drawn from one normal distribution and 10% from a second normal
distribution with the same mean but four times the standard deviation
[@Delacre:2019code]. That mixture has excess kurtosis 9.7, and is therefore
heavier-tailed than any input simulated here. They
caution that under skewness this robustness requires about 50 observations per group
for up to four groups and about 100 observations per group for more than four groups,
holding the sizes of $k-1$ groups fixed and perturbing only the last. Our design
perturbs all four group sizes simultaneously, so the two thresholds are not directly
comparable: at $\bar n=100$ every group exceeds 50 and Welch is robust, but already at
$\bar n=50$, with groups $(25,40,60,75)$, it is robust as well. The present results
are therefore consistent with their recommendation without being more demanding than
it, and they identify the one setting in which fixed Welch is not sufficient: skewness
2 with excess kurtosis 6.6 at a smallest group of 5 or 10 observations.

The Levene-gated mean route `L` leaves Bradley's bounds in six cells, all in panel (D)
and all at small samples: 9.87%, 9.21%, 9.51% and 9.96% at $\bar n=10$ for the
$(0,0)$, $(0.5,1)$, $(1,1.6)$ and $(2,6.6)$ inputs, and 7.73% and 8.88% at
$\bar n=20$ for $(1,1.6)$ and $(2,6.6)$. It is inside the bounds for every input from
$\bar n=50$ on. The reason is visible in its own route split: in the balanced
unequal-SD design the Levene gate still selects Fisher in 57.33% of replications at
$n=10$ and in 22.53% at $n=20$, but in 0.32% at $n=50$ and in 0.00% at $n=100$. The
Levene mean route thus improves on fixed Fisher/Student in the hardest cells without
reaching Bradley's bounds there: where the variance pretest is underpowered it passes
heteroscedastic data to Fisher and inherits Fisher's bias, and where it is powerful it
selects Welch and becomes redundant with a fixed Welch default
[@Moser:1992; @Zimmerman:2004; @Rochon:2012].

The default `SW+L` inherits both liabilities, and this is the honest price of keeping
Fisher's ANOVA and Student's t-test reachable by automated gating. In the two
symmetric columns of panel (D) — where, by rule 2, the reported rate is a Type I error
rate for both estimands, so no change of hypothesis can excuse it — `SW+L` leaves
Bradley's bounds in six of eight cells: 10.14% and 8.22% at $\bar n=10$, 7.83% at
$\bar n=20$, 7.58% at $\bar n=50$, and 8.07% and 7.76% at $\bar n=100$. Two unrelated
mechanisms act at opposite ends of the sample-size range. At $\bar n=10$ and
$\bar n=20$ the Levene gate is too weak: it routes 45.53% and 19.62% of replications
to Fisher's ANOVA, whose rate in this design is 13%, which is why `SW+L` (10.14%) is
worse there than `SW` (6.47%) and than `W` (5.96%). At $\bar n=50$ and $\bar n=100$
the Levene gate has become irrelevant, the Shapiro--Wilk gate routes 45.04% and 72.47%
of replications to the rank branch, and `SW+L` coincides with `SW` and inherits the
rank test's own sensitivity to unequal variances at unequal group sizes
[@Zimmerman:2004]. Fixed Welch stays inside Bradley's bounds in all eight of these
cells, between 3.96% and 5.96%.

## Type II error (power) simulations

The power simulation uses the same five input distributions and adds ordered location
shifts across the four groups. It uses balanced groups with equal SD, so $n_i=n$ and
$\mathrm{SD}_i=1$ for $i=1,\ldots,4$, and group means are shifted by
$0, 0.25, 0.50, 0.75$ SD. Rejection rates are shown for $n=10$ to $n=100$. The
question the figure answers is what the full `SW+L` default costs relative to `SW`,
that is what the second gate contributes once the shape gate has acted.

```{r route1-power, echo=FALSE, fig.height=7.8, fig.cap="Route 1 power simulation with Fleishman input distributions. (A) Input distributions with group mean and median reference lines. (B) Simulated rejection rates for the six testing strategies."}
knitr::include_graphics(
 "figures/fleishman_4groups_power.png"
)
```

Under the normal input the six strategies are indistinguishable at every $n$: at
$n=10$ they lie between 23.04% (`KW`) and 26.36% (`SW+L`), at $n=20$ between 48.35%
and 51.54%, and at $n=50$ between 90.63% and 92.06%. As skewness and tail weight
increase, the rank branch gains: at $n=10$ under the $(2,6.6)$ input `KW` rejects
43.00% against 29.62% for `F` and 31.62% for `W`, and at $n=20$ 78.46% against 54.19%
and 55.75%. Because the Shapiro--Wilk gate routes an increasing share of these
replications to the rank branch, `SW` and `SW+L` follow `KW` rather than the
mean-based tests in the skewed columns; the inset route percentages make that transfer
explicit. At $n=100$ all strategies exceed 99.7% and the comparison is uninformative.

The isolated contribution of the Levene gate is negligible over the whole power grid.
`SW+L` is never below `SW` by more than 0.002 percentage points, exceeds it by at most
2.25 percentage points, and the median difference across the twenty
distribution-by-sample-size cells is 0.004 percentage points. The second gate
therefore changes which mean-based test is reported, and the route display, but not
whether the omnibus null is rejected.

## What the simulations establish

Read strictly as an equal-means procedure, Route 1 is best served by
`group_test = "welch"`: fixed Welch is the only mean-based strategy that stays within
Bradley's bounds in all but two of the 100 equal-means cells, it loses no measurable
power where its assumptions are met, and its algebraic distance from Student's t-test
and Fisher's ANOVA under equal variances is $O(n^{-1})$. Read as a procedure that is
allowed to use input shape as a method switch, `SW` is the simpler automatic route,
and the default `SW+L` buys the same switch while preserving the Fisher/Student branch
of the general linear model when its two displayed assumptions pass. `group_test =
"rank"` is the fixed rank-based choice, but under skewed unequal-SD inputs it answers a
different question.

The simulations therefore do not show that `SW+L` is the best equal-means procedure.
They show its price and its benefit. The price is paid twice in the unbalanced design
whose smallest group carries the largest SD: under skewness the reported rate is
increasingly that of a rank test after the hypothesis has changed, and even under
symmetry the procedure is Type I non-robust there, through the weak variance gate at
small samples and through the rank branch at large ones. The benefit is that the
switch is visible and diagnostic-driven: Fisher's ANOVA or Student's t-test is used
only when both displayed general-linear-model assumptions pass, a Welch variant when
only homoscedasticity fails, and a rank test when residual normality fails, with the
route percentages in the figures documenting each of these decisions directly.
