---
title: "The $n_i > 50$ residual-normality bypass: scope, evidence, and boundary"
subtitle: "Working note for the visStatistics Discussion section"
date: "2026-06-01"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 4.5)
suppressMessages(library(visStatistics))
```

# The justification of the $n>50$ gate 

For grouped numeric responses (Route 1), `visstat()` applies a residual
normality test (Shapiro--Wilk) only when at least one group has
$n_i \le 50$. When **all** group sizes exceed 50, the residual-normality
test is bypassed: routing proceeds straight to the variance gate and then to
a mean-based test. Verified in `R/visstat_core.R` (`all(counts_per_level > 50)`).
The CLT [@Lumley:2002] is the **motivation** for a finite cutoff, not the
**evidence** that 50 suffices — the convergence rate of the group means to normality depends on skewness, so
the cutoff is a pragmatic choice the simulations above calibrate.


The bypass **fires before** the homoscedasticity (Levene) gate and before the
two-vs-more-groups split. At the moment it decides, it does **not** yet know
which of four mean-based tests will run. Its justification therefore cannot
rest on a single source — it must cover a 2×2 of branches.

Here is an overview of the literature regarding threshold values in the four branches: 

| | $k = 2$ | $k > 2$ |
|---|---|---|
| **equal variance** (Student / Fisher) | Blanca et al. (2017) — Student is the two-group special case | Blanca et al. (2017) |
| **unequal variance** (Welch-$t$ / Welch-ANOVA) | Fagerland (2012) | Delacre et al. (2019) |

- **Blanca et al. (2017)** — Fisher one-way ANOVA: Type I error within
  Bradley's bounds across skewness up to 2, kurtosis up to 6, group sizes
  5–100, *equal variances*.
- **Fagerland (2012)** — Welch's $t$, unequal variances, two groups:
  $n \approx 50$ holds the 5% rate across a wide skewness range; only heavy
  skew needs $\approx 200$.
- **Delacre et al. (2019)** — Welch's one-way ANOVA, unequal variances,
  $k > 2$: $\approx 50$ per group suffices, except for $k > 4$ under
  pronounced skew, where $\approx 100$ per group is needed. *(page to
  re-verify before manuscript use.)*



# An honest test case: equal means, different medians

To probe the bypass, build $k = 5$ lognormal groups with
**equal population means** but **unequal variance and skew**. For a lognormal,
$\text{mean} = e^{\mu + \sigma^2/2}$, so fixing $\mu_i = M - \sigma_i^2/2$
holds every group mean at $e^{M}$ while spread and skew rise with $\sigma_i$.
Side effect: medians $= e^{\mu_i}$ **differ**.

Under equal means, **every rejection by a mean-based test is a false
positive** — it cannot be faked by a built-in mean difference.

```{r construction}
sdlog   <- c(A = 0.3, B = 0.5, C = 0.7, D = 0.9, E = 1.1)  # rising spread/skew
M       <- 1                                               # every group mean = exp(M)
meanlog <- M - sdlog^2 / 2                                 # equalise the means

data.frame(
  group           = names(sdlog),
  pop_mean        = round(exp(meanlog + sdlog^2/2), 3),    # all equal
  pop_median      = round(exp(meanlog), 3),                # differ
  pop_sd_rank     = rank(sdlog)
)
```

```{r helpers}
make_data <- function(n) {
  grp <- factor(rep(names(n), times = n))
  val <- unlist(lapply(names(n), function(k) rlnorm(n[k], meanlog[k], sdlog[k])))
  list(grp = grp, val = val)
}

# pull the test visstat() actually ran (method + p) from its result object
visstat_test <- function(res) {
  for (el in res)
    if (is.list(el) && !is.null(el$method) && !is.null(el$p.value))
      return(list(method = trimws(el$method), p = el$p.value))
  list(method = "(not found)", p = NA_real_)
}
```

# What visstat() does, across three size regimes

Each regime keeps the same structure; only the group sizes change. The plots
are the default `visstat()` panels. The printed block contrasts the test
`visstat()` chose with Kruskal--Wallis.

```{r cases, fig.show='hold', results='hold'}
welch_case <- function(n, label) {
  set.seed(1)
  d   <- make_data(n)
  res <- visstat(d$grp, d$val)
  vt  <- visstat_test(res)
  kw  <- kruskal.test(d$val ~ d$grp)
  cat("===========", label, "===========\n")
  cat("means  :", paste(round(tapply(d$val, d$grp, mean),   2), collapse = " / "), "\n")
  cat("medians:", paste(round(tapply(d$val, d$grp, median), 2), collapse = " / "), "\n")
  cat(sprintf("visstat()  %-48s p = %.4g\n", vt$method, vt$p))
  cat(sprintf("Kruskal    %-48s p = %.4g\n\n", kw$method, kw$p.value))
}

welch_case(c(A = 55,  B = 65,  C = 75,  D = 85,  E = 95),  "n in (50,100)")
welch_case(c(A = 110, B = 130, C = 150, D = 170, E = 190), "n > 100")
welch_case(c(A = 510, B = 530, C = 550, D = 570, E = 590), "n > 500")
```

**Reading it:** means are equal, so `visstat()` (Welch's ANOVA) correctly does
*not* reject. Medians differ, so Kruskal--Wallis correctly does. The divergence
is a property of the question each test asks (mean vs. rank/location), not a
bug — and it is untouched by sample size.

# Does raising the threshold to 100 fix anything?

The single-draw view shows correctness on average. The real cost is the
**false-positive rate** under equal means. Bradley's liberal robustness bound
for $\alpha = .05$ is $[.025, .075]$.

```{r typeI}
welch_typeI <- function(n, nrep = 2000, alpha = 0.05) {
  grp <- factor(rep(names(n), times = n))
  set.seed(1)
  mean(replicate(nrep, {
    val <- unlist(lapply(names(n), function(k) rlnorm(n[k], meanlog[k], sdlog[k])))
    oneway.test(val ~ grp, var.equal = FALSE)$p.value < alpha
  }))
}

data.frame(
  regime       = c("n in (50,100)", "n > 100", "n > 500"),
  welch_typeI  = c(
    welch_typeI(c(A = 55,  B = 65,  C = 75,  D = 85,  E = 95)),
    welch_typeI(c(A = 110, B = 130, C = 150, D = 170, E = 190)),
    welch_typeI(c(A = 510, B = 530, C = 550, D = 570, E = 590))
  )
)
```

For this $k = 5$ skewed, heteroscedastic corner: the false-positive rate is
inflated at $n \in (50,100)$, only **borderline** at $n > 100$ (sitting on
Bradley's ceiling), and comfortably controlled only well past 100 per group.
So a blanket "raise the cutoff to 100" does **not** cleanly fix it — matching
Delacre et al. (2019). A single group-size threshold cannot cover the
$k > 4$ skewed-heteroscedastic corner.

# Consequence for the package

- Keep the bypass and the cutoff at 50: clearly supported for the cases the
  package targets (the 2×2 above).
- Do **not** chase the corner with a $k$-dependent threshold: it cannot be
  made clean, and the bypass deliberately favours not switching to rank-based
  tests over negligible large-sample deviations.
- Instead, rely on the package's existing safeguards: every test is reported
  with an **effect size**, and the **residual diagnostics are displayed**, so a
  significant mean-based result paired with a small effect size stays visible.

# Proposed Discussion paragraph (replaces current L898)

> It also depends on which test the routing will reach. Holding the
> false-positive rate to its $\alpha = 5\%$ target, simulation evidence
> supports the threshold of 50 across all of these branches. For equal
> variances---Fisher's one-way ANOVA, with Student's t-test as its two-group
> special case---a Monte Carlo study kept the Type I error within Bradley's
> bounds in every condition examined: skewness up to 2, kurtosis up to 6, and
> group sizes from 5 to 100 [@Blanca:2017]. For unequal variances with two
> groups---Welch's t-test---group sizes near 50 hold the 5% rate across a wide
> range of skewness, and only heavily skewed distributions call for
> substantially larger samples, on the order of 200 [@Fagerland:2012]. For
> unequal variances with more than two groups---Welch's one-way ANOVA---about
> 50 observations per group suffice, except when more than four groups are
> compared under pronounced skew, where roughly 100 per group are required
> [@Delacre:2019].

**Before manuscript use:** (1) re-verify the Delacre page for the
"$\approx 100$ for $k > 4$" claim; (2) ensure `@Blanca:2017` is in
`REFERENCES.bib` (author-managed).
