---
title: "Route 1 Assumption Plot Notes"
output: html_document
---

# Purpose

This note summarises the current design decision for changing the Route 1
assumption-diagnostic panel in `visStatistics`.

# Core Distinction

- Normality checks ask whether model errors have the expected distributional
  shape.
- Homoscedasticity checks ask whether model residual spread is constant.
- Therefore, normality and homoscedasticity do not have to use the same
  residual scale.

# Normality Diagnostics

Normality is checked on raw model residuals:

```{r, eval=FALSE}
fit <- lm(y ~ group)
e <- residuals(fit)
shapiro.test(e)
nortest::ad.test(e)
```

Rationale:

- formal p-values use the residual scale of the fitted model;
- common scaling would not change the normality-test p-values;
- Route 1 plots use `e / sigma(fit)` only for visual comparison.

# Homoscedasticity Diagnostics

Variance tests should use raw model residuals:

```{r, eval=FALSE}
fit <- lm(y ~ group)
e <- residuals(fit)

bartlett.test(e ~ group)
summary(aov(abs(e) ~ group))
```

Rationale:

- homoscedasticity concerns residual spread;
- spread is the target and should not be removed by standardisation;
- Bartlett and mean-centred Levene-type tests are defined on residual
  deviations or variances;
- using raw residuals keeps the p-values aligned with the model-scale variance
  assumption.

# Plot Design

Recommended Route 1 diagnostic panel:

- Panel 1: histogram and normal-density overlay of `e / sigma(fit)`.
- Panel 2: Q--Q plot of `e / sigma(fit)`.
- Panel 3: residual spread by group using `abs(e / sigma(fit))`.

Panel 3 uses:

```{r, eval=FALSE}
abs(residuals(fit) / sigma(fit))
```

Rationale:

- dividing by `sigma(fit)` gives a readable standard-error scale;
- the same scaling is applied to all observations;
- relative group spread is preserved;
- unlike `rstandard()`, this does not introduce leverage-dependent scaling.

# Proposed Labels

- Panel 1 title: `Histogram and Normal Density`
- Panel 1 x-axis: `z-transformed residuals`
- Panel 2 title: `Normal Q-Q Plot`
- Panel 2 y-axis: `z-transformed residuals`
- Panel 3 title: `Absolute z-residual spread`
- Panel 3 y-axis: `|z_i|`

# Outer Title

Keep the current two-line structure:

```{r, eval=FALSE}
Linear model assumptions: Shapiro-Wilk p = ... | Anderson-Darling p = ...
Levene p = ... | Bartlett p = ...
```

The Levene p-value should refer to the mean-centred Levene-type test on raw
model residuals. Bartlett should also be calculated on raw model residuals.

# Implementation Decision

The package implementation now uses the mean-centred `levene.test()` on raw
model residuals.
