---
title: "Unified Framework of t-test, ANOVA, and Linear Models under the General Linear Model"
author: "Sabine Schilling"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    mathjax: default
# bibliography: references.bib
#csl: apa.csl
---

# Introduction

The **general linear model (GLM)** serves as the unifying mathematical structure for many classical statistical techniques, including the **t-test**, **one-way ANOVA**, and **linear regression**, whether simple or multiple. All these methods share the same assumptions and estimation framework. This document demonstrates this connection, proves the equivalence \( t^2 = F \), and interprets model coefficients.

# The General Linear Model

A general linear model is given by:

$$
\mathbf{y} = \mathbf{X}\boldsymbol\beta + \boldsymbol\varepsilon, \quad \boldsymbol\varepsilon \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}_n)
$$

Where:
- \( \mathbf{y} \in \mathbb{R}^n \): response vector  
- \( \mathbf{X} \in \mathbb{R}^{n \times p} \): model matrix  
- \( \boldsymbol\beta \in \mathbb{R}^p \): regression coefficients  
- \( \boldsymbol\varepsilon \): homoscedastic, independent, normally distributed errors  

This framework applies whether \( \mathbf{X} \) encodes group indicators (as in t-tests or ANOVA) or continuous predictors (as in regression). See [Faraway, 2014](#faraway2014) and [Fox, 2015](#fox2015).

# Shared Assumptions

All the following methods are GLMs under the same assumptions:
1. **Linearity**: The expected value of the response is a linear combination of predictors.
2. **Independence**: Observations are independent.
3. **Homoscedasticity**: Constant variance \( \operatorname{Var}(\varepsilon_i) = \sigma^2 \).
4. **Normality**: \( \varepsilon_i \sim \mathcal{N}(0, \sigma^2) \) for exact inference.

# Equivalence of t-test, ANOVA, and `lm()`

## Two-sample t-test as a linear model

```{r}
set.seed(42)
group <- factor(rep(c("A", "B"), each = 10))
y <- c(rnorm(10, mean = 5), rnorm(10, mean = 7))

t.test(y ~ group) # Welch's t-test

model_t <- lm(y ~ group)
summary(model_t)

anova(model_t)
```

- The coefficient for `groupB` is the **mean difference**.
- The square of the `t` statistic from `summary(model_t)` equals the `F` from `anova()`.

```{r}
t_val <- summary(model_t)$coefficients[2, "t value"]
f_val <- anova(model_t)[1, "F value"]
c(t_squared = t_val^2, F = f_val)
```

## One-way ANOVA as a GLM

```{r}
group3 <- factor(rep(c("A", "B", "C"), each = 10))
y3 <- c(rnorm(10, 5), rnorm(10, 7), rnorm(10, 6))

model_aov <- aov(y3 ~ group3)
summary(model_aov)

model_lm <- lm(y3 ~ group3)
anova(model_lm)
```

## Multiple regression as GLM

```{r}
x1 <- rnorm(30)
x2 <- rnorm(30)
y_multi <- 2 + 1.5 * x1 - 0.7 * x2 + rnorm(30)

model_multi <- lm(y_multi ~ x1 + x2)
summary(model_multi)
```

# Mathematical Identity: \( t^2 = F \)

For testing a single coefficient \( \beta_j \) in linear regression:

$$
t_j = \frac{\hat\beta_j}{\operatorname{SE}(\hat\beta_j)} \quad \text{and} \quad F = t_j^2
$$

This holds for any one-parameter linear contrast, under the null hypothesis.

**Proof**:  
If \( H_0: \beta_j = 0 \), then:

- Numerator SS: \( \text{SS}_{\text{reg},j} = \hat\beta_j^2 / \operatorname{Var}(\hat\beta_j) \)
- Denominator MSE: \( \hat\sigma^2 = \text{RSS} / (n - p) \)

Then:

$$
F = \frac{\hat\beta_j^2}{\operatorname{Var}(\hat\beta_j)} / \hat\sigma^2 = t_j^2
$$

# Interpretation of Coefficients

In the GLM:

- \( \beta_0 \): the expected value of \( y \) when all predictors are 0.  
  - In group comparisons: mean of reference group  
  - In regression: intercept at \( x_1 = x_2 = \dots = 0 \)

- \( \beta_1, \beta_2, \dots \): expected change in \( y \) per unit increase in predictor, holding others constant.

In t-tests and ANOVA:
- Coefficients represent **group means** or **mean differences**, interpretable as regression slopes on dummy variables.

# Conclusion

The t-test, ANOVA, and regression models are all **specific cases** of the **general linear model**, differing only in how the predictor matrix \( \mathbf{X} \) is constructed. They share assumptions, estimation, and inferential logic.

# References

```{r, echo = FALSE, results = "asis"}
# Generate references from .bib
```

---


```bibtex
@book{faraway2014,
  title     = {Linear Models with R},
  author    = {Julian J. Faraway},
  year      = {2014},
  publisher = {CRC Press},
  edition   = {2},
  doi       = {10.1201/b17107}
}

@book{fox2015,
  title     = {Applied Regression Analysis and Generalized Linear Models},
  author    = {John Fox},
  year      = {2015},
  edition   = {3},
  publisher = {SAGE Publications},
  doi       = {10.4135/9781483393126}
}
```